How Slow do my Transient Typechecks Go?

### 2019-08-17 paper.tex
Reworded up to section 4

Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.

@@

\paragraph{Moth: A Grace Implementation}
Moth is a virtual machine for Grace that we are developing as a
research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
based on the Truffle and Graal toolchain, Moth is a self-optimising
AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
The key idea is that an AST rewrites itself based on a program's run-time values
to reflect the minimal set of operations needed to execute the program
correctly. The rewritten AST is then compiled into efficient machine
code. This rewriting often depends on the dynamic types of the
objects involved. In the simplest case, a ``self'' call (when one method
on an object invokes a second method on the exact same object) will
always result in executing the second method defined in same concrete
class as the first method. The second method can then be inlined into
the first method, avoiding the overheads of a machine-level subroutine
invocation and of an object-oriented dynamic dispatch.

Moth relies on a number of standard techniques for optimising
object-oriented programs.
``Shapes'' \citep{woss2014object} capture information about objects'
structure and (run time) 
field types, allowing a just-in-time compiler to
represent objects in memory similarly to C structs and, consequently,
can generate highly efficient code.
``Polymorphic inline caches''
\citep{Hoelzle:91:PIC} uses object shapes to cache the results of
method lookups, avoiding expensive class hierarchy searches or
indirect jumps through virtual method tables. 
For languages built on the Truffle framework,
Graal comes with  additional support for partial evaluation,
which enables efficient native code generation for
Truffle interpreters \citep{Wurthinger:2017:PPE}.
\RR{This last sentence seems out of place, does it belong here?}

@@

In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
performance of transient typechecks using the \AWFY
benchmark suite \cite{Marr2016}, plus other
benchmarks from the gradual-typing literature.
The goal was to complement our benchmarks with additional ones that are
used for similar experiments and can be ported to Grace.
To this end, we surveyed a number of papers
\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected benchmarks that have been used by multiple papers.
Some of these benchmarks overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviorally equivalent,
we chose the \AWFY versions since we already used them to
establish the performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
Table~\ref{tab:gradual-benchmarks}.

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth runs programs with
transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mixed typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, the so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark, where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per-module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et al.\ modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
a maximum of 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

To investigate Moth's support for transient typechecks further, we
carried out a (small?) Takikawa-Vitousek evaluation using (some of?)
our existing benchmarks.

\textbf{!!!KJX: Need to update this paragraph with what we did this time around!!!}
To account for the complex warmup behavior
of modern systems\citep{Barrett:2017:VMW} as well as
the non-determinism caused by \eg garbage collection and cache effects,
we run each benchmark for \textbf{$/$NumIterationsAll} iterations in the same
VM invocation.\footnote{
For the Higgs VM, we only use \textbf{$/$NumIterationsHiggs} iterations,
because of its lower performance.
This is sufficient since Higgs's compilation approach induces less variation
and leads to more stable measurements.}
Afterwards, we inspected the run-time plots over the iterations
and manually determined a cutoff of \textbf{$/$WarmupCutOff} iterations for warmup,
\ie, we discard iterations with signs of compilation.
As a result, we use a large number of data points to compute the average,
but outliers, caused by \eg garbage collection, remain visible in the plots.
% In this work, we do not consider startup performance,
% because we want to assess the impact of gradual type checks
% on the best possible performance.
All reported averages use the geometric mean since they aggregate
ratios.

\textbf{
% Yuria
%  - Ubuntu 16.04.4, Kernel 3.13
%  - 24 hyperthreads
%  - Intel Xeon E5-2620 v3 2.40GHz
% Graal 0.33 Feb. 2018
%
% TODO: update these details when we have time... not a rejection criterion,
% but needs to be correct in the final version
All experiments were executed on a machine running Ubuntu Linux 16.04.4,
with Kernel 3.13.
The machine has two Intel Xeon E5-2620 v3 2.40GHz,
with 6 cores each, for a total of 24 hyperthreads.
We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results was done with R 3.4.1,
and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
Our experimental setup is available online to enable reproductions.\footnote{
\url{https://github.com/gracelang/moth-benchmarks}}
}

@@

The high-performance computing community also has been investigating how
tools and visualization can help developers to utilize their systems
more efficiently \citep{Papenhausen:2016:IVT,daSilva:2019:PSV}.
Their focus is typically on parallelisation opportunities,
guided by runtime feedback, cost models, or heuristics. 
The large body of work\citep{Isaacs:2014:PerfViz} uses various approaches,
though, we are not aware of work has used an approach as ours.
One may argue, this is largely a question of practicality.\sm{
i put this in to hint at issues with usability.}

>>>>

\paragraph{Moth's Transient Type Checking}
Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.

@@

\paragraph{Moth's Optimisation}
We are developing Moth as a
research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
based on the Truffle and Graal toolchain, Moth is a self-optimising
AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
The key idea is that an AST rewrites itself based on a program's run time values
to reflect the minimal set of operations needed to execute the program
correctly. The rewritten AST is then compiled into efficient machine
code. This rewriting often depends on the dynamic types of the
objects involved. In the simplest case, a ``self'' call (when one method
on an object invokes a second method on the exact same object) will
always result in executing the exact same method. Thus the called method can be inlined into
the callee, avoiding overhead of a machine-level subroutine
invocation and an object-oriented dynamic dispatch.

Moth relies on a number of standard techniques for optimising
object-oriented programs.
``Shapes'' \citep{woss2014object} capture information about objects'
structure and (run time) 
field types, allowing a just-in-time compiler to
represent objects in memory similarly to C structs and, consequently,
can generate highly efficient code.
``Polymorphic inline caches''
\citep{Hoelzle:91:PIC} use object shapes to cache the results of
method lookups, avoiding expensive class hierarchy searches or
indirect jumps through virtual method tables. 
Since Moth is built on the Truffle framework,
Graal comes with  additional support for partial evaluation,
which enables efficient native code generation for
Truffle interpreters \citep{Wurthinger:2017:PPE}.
\RR{This last sentence seems out of place, does it belong here?}

@@

In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
performance of transient type checks using 21 benchmarks.
Including the 14 benchmarks of the \AWFY
benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks. We also used benchmarks taken from the gradual-typing literature.
Our goal was to complement our benchmarks with additional ones that are used for similar experiments and can be ported to Grace.
To this end, we surveyed a number of papers
\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected 9, benchmarks that have been used by multiple papers.
Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviourally equivalent,
we chose the \AWFY versions since we already used them to
establish the performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
Table~\ref{tab:gradual-benchmarks}.

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we ensured that each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth's transient type checking has an overhead of 5\% (min.\ -13\%,
max. 79\%) compared to Moth's peak performance with type checking disabled.

We did not measure programs that mixed typed and untyped code because
with our implementation technique we expected fully typed programs to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, the so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each
benchmark, where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per-module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose whether each individual declaration should be type-checked: so here $N$ is the number of type annotations in the program,
so checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
a maximum of 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

To investigate Moth's support for transient type checks further, we carried out a similar approach to Takikawa-Vitousek evaluation: for each benchmark we generated 100 partially typed versions (or less if the benchmark was too small). We did an even split so that for each $i \ge 1$ and $i$ < $N$ (where $N$ is the total number of type annotation), we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations where generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark, (or 97 in the case of Storage, since it only has 10 type annotations).

As we did in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems\citep{Barrett:2017:VMW} as well as
the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and used cut off of 350 iterations for warmup, \ie we discard the first 350 iterations, since they showed signs of compilation.

Though though outliers, caused by \eg garbage collection, remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).

All experiments were executed on a Windows 10 machine
running Fedora 29 within Windows Subsystems for Linux (version 1). The machine has

one Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16.
To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, is available online.\footnote{
\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

@@

The high-performance computing community also has been investigating how
tools and visualization can help developers to utilize their systems
more efficiently \citep{Papenhausen:2016:IVT,daSilva:2019:PSV}.
Their focus is typically on parallelisation opportunities,
guided by runtime feedback, cost models, or heuristics. 
The large body of work\citep{Isaacs:2014:PerfViz} uses various approaches,
though, we are not aware of work that has used an approach similar to ours.
One may argue, this is largely a question of practicality.\sm{
i put this in to hint at issues with usability.}
